---
title: "Bootstrap the Super Bowl! Comparing QBs (Yards per Completion) and RBs (Yards per Carry"
output: html_document
date: "2024-05-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Motivate the topic.

Compare QBs and RBs - is it all about the average?  yards per completion, yards per carry

Do big plays matter?

Bootstrapping helps us estimate variability in our estimated statistic.

# State the Learning Objectives.

# Introduce the data.

# What is bootstrapping?

Instruction goes here

# Example - QB comparison...

# Student Challenge!

Do the same thing for McCaffrey (49ers) and Pacheco (Chiefs)...

# Conclusion / Recap

```{r}
library(nflfastR)
library(tidyverse)
```

```{r}
pbp <- nflfastR::load_pbp(2023)
```

```{r}
pbp %>% 
  filter(passer_player_name %in% c("P.Mahomes", "B.Purdy")) %>% 
  group_by(passer_player_name) %>% 
  count()
```

```{r}
pbp %>% 
  filter(passer_player_name %in% c("P.Mahomes", "B.Purdy"),
         play_type == "pass") %>%
  filter(str_detect(game_id, "2023_22")) %>% 
  group_by(passer_player_name) %>% 
  summarize(mean = mean(passing_yards, na.rm = TRUE),
            n = n)
```

```{r}
sb_qb_df <- pbp %>% 
  filter(passer_player_name %in% c("P.Mahomes", "B.Purdy"),
         play_type == "pass",
         complete_pass == 1) %>%
  filter(str_detect(game_id, "2023_22"))
```

```{r}
sb_qb_df %>% 
  ggplot(aes(x = passing_yards)) + 
  geom_histogram() + 
  facet_wrap(~passer_player_name, ncol = 1)
```

```{r}
# Compute sample means
sb_qb_df %>% 
  group_by(passer_player_name) %>% 
  summarize(mean = mean(passing_yards),
            sd = sd(passing_yards),
            n = n())
```

```{r}
# Construct bootstrap CIs for each QB
mahomes <- sb_qb_df %>% 
  filter(passer_player_name == "P.Mahomes") %>% 
  pull(passing_yards)
purdy <- sb_qb_df %>% 
  filter(passer_player_name == "B.Purdy") %>% 
  pull(passing_yards)
mahomes_boot_dist <- replicate(n = 1000, sample(mahomes, replace = TRUE) %>% mean())
purdy_boot_dist <- replicate(n = 1000, sample(purdy, replace = TRUE) %>% mean())

quantile(mahomes_boot_dist, c(0.025,0.975))
quantile(purdy_boot_dist, c(0.025,0.975))

t.test(mahomes, purdy)
```

# Running Back Comparison

```{r}
sb_rb_df <- pbp %>% 
  filter(rusher_player_name %in% c("C.McCaffrey", "I.Pacheco"),
         play_type == "run") %>%
  filter(str_detect(game_id, "2023_22"))
```

